---
title: "Network Science: From Theory to Biological Applications"
subtitle: "BC204 Hands-on Workshop · Spring 2026"
author: "Savvas Paragkamian"
date: today
format:
  html:
    toc: true
    toc-float: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    theme: flatly
    number-sections: true
    fig-width: 8
    fig-height: 5
    embed-resources: true
execute:
  warning: false
  message: false
bibliography: ../Bibliography/Network Science Workshop 2018.bib
---

```{r}
#| label: setup
#| code-fold: true

required <- c("igraph", "ggraph", "tidygraph", "poweRlaw", "tidyverse", "knitr")
missing  <- required[!required %in% installed.packages()[, "Package"]]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")

library(igraph)
library(ggraph)
library(tidygraph)
library(poweRlaw)
library(tidyverse)
library(knitr)

theme_net <- function() {
  theme_bw() +
    theme(panel.grid.minor = element_blank(),
          strip.background = element_blank())
}
```

---

# Part 0 · Data Structures and Network Types `[~30 min]`

> **Goal**: Build the vocabulary and the R objects we will use for the rest of the day. Every network analysis starts here.

## Complicated ≠ Complex

These words are used interchangeably but they are not the same, and the distinction determines what kind of model is appropriate.

A **complicated** system has many parts, but the parts compose trivially. A mechanical watch has hundreds of cogs and springs, each with a specific job. Take it apart, document every piece, and you have a complete description. Behaviour is the sum of parts. Complicated systems need detailed blueprints.

A **complex** system has parts that *interact*, with feedbacks and nonlinearities. The interesting behaviour lives in the couplings, not the components. A three-species food web can show oscillations and regime shifts that no single species possesses. Complex systems often admit strikingly simple mathematical descriptions — because the richness is generated by a few interaction rules, not by idiosyncratic detail.

Start with the simplest possible population model. **Malthusian growth** assumes a population grows proportionally to its current size — no resource limits, no interactions:

$$N_{t+1} = r \cdot N_t \qquad \Longrightarrow \qquad N_t = N_0 \cdot r^t$$

```{r}
#| label: malthusian
#| fig-cap: "Malthusian dynamics: three growth rates, exponential trajectories. No interactions — the system is complicated only if N₀ is large, but never complex."

malthus <- function(r, n0 = 1, steps = 20) {
  tibble(t = 0:steps, N = n0 * r^(0:steps), r = as.character(r))
}

map_dfr(c(0.8, 1.0, 1.2), malthus) |>
  ggplot(aes(x = t, y = N, colour = r)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c("0.8" = "steelblue", "1" = "grey50", "1.2" = "coral2"),
                      labels = c("r = 0.8  (decline)", "r = 1.0  (steady)", "r = 1.2  (growth)")) +
  labs(x = "Time step", y = "Population N", colour = NULL,
       title = "Malthusian (exponential) growth") +
  theme_net()
```

Malthusian dynamics is **not complex** — it is entirely predictable. The behaviour of the whole is the behaviour of one individual, scaled. No emergence is possible.

Adding a single interaction term — **density dependence**, the idea that crowding limits growth — transforms the model entirely:

$$x_{t+1} = r \cdot x_t (1 - x_t)$$

The $(1 - x_t)$ term makes each individual interact with the rest of the population. One interaction rule. One parameter.

```{r}
#| label: logistic-map
#| fig-cap: "Logistic map bifurcation diagram: stable fixed point → period doubling → chaos. The only change from the Malthusian model is the (1 − x) term."

logistic_attractor <- function(r, n_burn = 500, n_keep = 300, x0 = 0.5) {
  x <- x0
  for (i in seq_len(n_burn))  x <- r * x * (1 - x)
  xs <- numeric(n_keep)
  for (i in seq_len(n_keep)) { x <- r * x * (1 - x); xs[i] <- x }
  xs
}

r_vals <- seq(2.5, 4.0, length.out = 800)

map_dfr(r_vals, function(r) tibble(r = r, x = logistic_attractor(r))) |>
  ggplot(aes(x = r, y = x)) +
  geom_point(size = 0.05, alpha = 0.3, colour = "steelblue") +
  labs(x = "Growth rate r", y = "Long-run population x",
       title = "Logistic map bifurcation diagram") +
  theme_net()
```

Beyond $r \approx 3.57$ the orbit never settles. The system is **deterministic** — the same $r$ and $x_0$ always give the same trajectory — yet **unpredictable**: two trajectories starting at almost the same $x_0$ rapidly diverge.

```{r}
#| label: chaos-sensitivity
#| fig-cap: "Sensitivity to initial conditions at r = 3.9: two trajectories differing by 0.0001 are indistinguishable for ~15 steps, then diverge completely. This is the hallmark of deterministic chaos."

logistic_ts <- function(r, x0, steps = 60) {
  x <- numeric(steps + 1); x[1] <- x0
  for (i in seq_len(steps)) x[i + 1] <- r * x[i] * (1 - x[i])
  tibble(t = 0:steps, x = x)
}

r_chaos <- 3.9
bind_rows(
  logistic_ts(r_chaos, x0 = 0.5000) |> mutate(start = "x₀ = 0.5000"),
  logistic_ts(r_chaos, x0 = 0.5001) |> mutate(start = "x₀ = 0.5001")
) |>
  ggplot(aes(x = t, y = x, colour = start)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("steelblue", "coral2")) +
  labs(x = "Time step", y = "Population x", colour = "Initial condition",
       title  = "Deterministic chaos: two nearly identical initial conditions",
       subtitle = paste0("r = ", r_chaos, "  |  Δx₀ = 0.0001")) +
  theme_net()
```

Beyond ~3.57 the orbit never settles — deterministic chaos from a deterministic rule. **This is emergence**: qualitatively new behaviour that is not present in any single part.

A **cell** is both complicated *and* complex. Billions of molecular machines, each with specific evolved structure (complicated) interacting through feedback loops and signalling cascades that produce homeostasis, differentiation, and cell fate decisions (complex). This is why whole-cell modelling is hard in a way that aircraft engineering and magnetism are not: we cannot get away with just the blueprint, and we cannot get away with just the coarse-grained dynamics.

| System | Complicated | Complex | Implication |
|---|:---:|:---:|---|
| Mechanical watch | ✓ | — | Detailed blueprint is the right model |
| Logistic map / Ising model | — | ✓ | Simple equation captures everything |
| Epidemic, food web | — | ✓ | A few interaction rules suffice |
| Cell, brain, ecosystem | ✓ | ✓ | Neither alone works — genuinely hard |

**Networks are the natural language of complex systems.** They make interactions explicit: nodes are the parts, edges are the couplings. The same formalism applies to neurons, proteins, species, and — as we will see in Part 5 — artificial intelligence.

## The cell is not a bag of genes

JBS Haldane described the cell as a "bag of genes". We know differently now. The first things we encounter inside a cell are not genes but **molecular machines** — elaborate, evolved structures that do specific physical work: the ribosome translates mRNA, the spliceosome edits it, myosin pulls on actin, the nuclear pore complex controls everything entering or leaving the nucleus.

Some machines are well-modelled. Ribosome queueing along mRNA has been studied since the 1960s using the **TASEP** (Totally Asymmetric Simple Exclusion Process) — a particle-exclusion model on a graph — borrowing from statistical physics. Myosin has its own lineage stretching back to Huxley's 1957 cross-bridge theory.

Many others are strikingly under-modelled despite spectacular cryo-EM structures:

```{r}
#| label: modelling-gap

tribble(
  ~Machine,               ~Subunits, ~`Known structure`, ~`Mathematical models (approx.)`, ~`Key function`,
  "Ribosome",             "~80",     "Yes (Å resolution)", "Thousands (TASEP, stochastic)",  "mRNA translation",
  "Myosin II",            "~6",      "Yes",                "Hundreds (cross-bridge theory)",  "Muscle contraction",
  "Spliceosome",          ">100",    "Yes (cryo-EM)",      "~Tens",                           "pre-mRNA splicing",
  "Chaperonin (GroEL)",   "~14",     "Yes",                "~Tens",                           "protein folding",
  "V-ATPase",             "~30",     "Yes",                "~Tens",                           "proton pumping",
  "Nuclear pore complex", "~120",    "Yes (cryo-ET)",      "Very few",                        "nuclear transport"
) |> kable(caption = "Molecular machines: structural knowledge vs mathematical models (after Stumpf)")
```

The challenge is not obtaining structures — cryo-EM has transformed that. The hard step is translating **structure → dynamics → predictive mathematical model**. And fewer than 1% of published models ever reach shared repositories like BioModels.

The TASEP model of ribosome translation is a worked example: a **graph process** on a path graph (mRNA sites), where particles (ribosomes) hop forward if the next site is empty. A single rule generates three distinct emergent phases.

```{r}
#| label: tasep
#| fig-cap: "TASEP ribosome simulation: low entry rate (α=0.1) vs high entry rate (α=0.7). Same rule, qualitatively different density profiles — emergence from a simple interaction."

tasep_sim <- function(n_sites = 100, alpha = 0.3, beta = 0.8,
                      p = 1, n_steps = 40000, n_sample = 200) {
  lattice <- integer(n_sites)
  density <- numeric(n_sites)
  samples <- 0L

  for (step in seq_len(n_steps)) {
    # entry
    if (lattice[1] == 0 && runif(1) < alpha) lattice[1] <- 1L
    # bulk hops (right to left to avoid double moves)
    for (i in (n_sites - 1):1) {
      if (lattice[i] == 1L && lattice[i + 1] == 0L && runif(1) < p)
        { lattice[i + 1] <- 1L; lattice[i] <- 0L }
    }
    # exit
    if (lattice[n_sites] == 1L && runif(1) < beta) lattice[n_sites] <- 0L
    # sample after burn-in
    if (step > n_steps / 2) { density <- density + lattice; samples <- samples + 1L }
  }
  tibble(site = seq_len(n_sites), density = density / samples)
}

set.seed(42)
bind_rows(
  tasep_sim(alpha = 0.1, beta = 0.8) |> mutate(condition = "Low entry (α=0.1) — low-density phase"),
  tasep_sim(alpha = 0.7, beta = 0.8) |> mutate(condition = "High entry (α=0.7) — high-density phase")
) |>
  ggplot(aes(x = site, y = density, colour = condition)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("steelblue", "coral2")) +
  labs(x = "mRNA position (codon)", y = "Mean ribosome occupancy",
       title = "TASEP: ribosome density along mRNA",
       colour = NULL) +
  theme_net() +
  theme(legend.position = "bottom")
```

> **Q**: The spliceosome has over 100 subunits and we have near-atomic resolution structures — yet almost no mechanistic models. Is this a complicated problem, a complex one, or both? What would a network model of the spliceosome actually represent?

**Networks are the natural language of complex systems.** They make interactions explicit: nodes are the parts, edges are the couplings. The same formalism applies to neurons, proteins, species, and — as we will see in Part 5 — artificial intelligence.

## What is a graph?

A network is modelled mathematically as a **graph** $G(V, E)$: a set of $N$ nodes $V$ and a set of edges $E \subseteq V \times V$.

Four assumptions underpin this abstraction:

1. All nodes belong to the same variable type (e.g., all proteins, or all neurons).
2. All edges belong to the same variable type (e.g., all physical interactions).
3. Each edge connects exactly two nodes — a **binary relation**.
4. Both nodes and edges are time-independent (a static snapshot).

These assumptions are often violated in real biology. Knowing *when* they break tells you when to upgrade to a richer model (temporal, multilayer, hypergraph).

## Data structures for networks

A network can be stored in three equivalent ways:

| Structure | Best for |
|---|---|
| **Adjacency matrix** | Dense networks, linear algebra, visualisation as a heatmap |
| **Edge list** | Sparse networks, tidy data pipelines |
| Incidence matrix | Bipartite networks (two node types) |

### Adjacency matrix

Rows and columns are nodes; a cell is 1 if an edge exists (0 otherwise). For a **directed** graph the matrix is asymmetric — entry $(i, j) = 1$ means an edge *from* $i$ *to* $j$. For an **undirected** graph it is symmetric.

```{r}
#| label: adjacency-matrix

set.seed(7)
n_nodes <- 5
node_names <- LETTERS[1:n_nodes]

adj <- matrix(
  sample(c(0L, 1L), n_nodes^2, replace = TRUE, prob = c(0.5, 0.5)),
  nrow = n_nodes, ncol = n_nodes,
  dimnames = list(node_names, node_names)
)
diag(adj) <- 0L  # no self-loops

kable(adj, caption = "Adjacency matrix — rows are sources, columns are targets")
```

Convert to an igraph object and plot:

```{r}
#| label: adj-to-graph
#| fig-cap: "Graph built from the adjacency matrix"

g_adj <- graph_from_adjacency_matrix(adj, mode = "directed")
plot(g_adj, vertex.color = "steelblue", vertex.label.color = "white",
     edge.arrow.size = 0.6, main = "From adjacency matrix")
```

Visualise the same matrix as a heatmap — useful for spotting asymmetry (directed) or symmetry (undirected):

```{r}
#| label: adj-heatmap
#| fig-cap: "Adjacency matrix as a heatmap — asymmetry signals a directed network"

as.data.frame(adj) |>
  rownames_to_column("from") |>
  pivot_longer(-from, names_to = "to", values_to = "edge") |>
  ggplot(aes(x = to, y = fct_rev(from), fill = factor(edge))) +
  geom_tile(colour = "white") +
  scale_fill_manual(values = c("0" = "grey92", "1" = "steelblue"),
                    labels = c("absent", "present")) +
  scale_x_discrete(position = "top") +
  coord_fixed() +
  labs(x = "To", y = "From", fill = "Edge", title = "Adjacency matrix heatmap") +
  theme_net()
```

### Edge list

An edge list stores only the existing edges — one row per edge. It is more memory-efficient than a matrix for sparse networks (most real biological networks have density < 1 %).

```{r}
#| label: edge-list

el <- as.data.frame(which(adj == 1, arr.ind = TRUE)) |>
  transmute(from = node_names[row], to = node_names[col])

kable(el, caption = "Edge list derived from the same network")
```

Build a graph directly from the edge list:

```{r}
#| label: el-to-graph
#| fig-cap: "The same graph built from the edge list — identical result"

g_el <- graph_from_edgelist(as.matrix(el), directed = TRUE)
plot(g_el, vertex.color = "coral2", vertex.label.color = "white",
     edge.arrow.size = 0.6, main = "From edge list")
```

### Weighted edges

Edges can carry a numeric **weight** (interaction strength, distance, correlation). Weights extend binary graphs to continuous ones.

```{r}
#| label: weighted-edgelist

el_w <- el |> mutate(weight = round(runif(n(), 0.1, 1.0), 2))

kable(el_w, caption = "Weighted edge list")
```

## Network types

Real networks differ in four properties you must check before any analysis:

| Property | Values | Example |
|---|---|---|
| **Direction** | directed / undirected | synapses (directed); co-expression (undirected) |
| **Weight** | weighted / binary | interaction strength; presence/absence |
| **Sign** | signed / unsigned | activation (+) / inhibition (−) in gene regulation |
| **Simplicity** | simple / multigraph | single edge per pair; multiple experimental evidence |

```{r}
#| label: network-type-checklist

g_w <- graph_from_data_frame(el_w, directed = TRUE)
E(g_w)$weight <- el_w$weight

tibble(
  Property  = c("Directed", "Weighted", "Simple", "Has loops"),
  Value     = c(is_directed(g_w), is_weighted(g_w),
                is_simple(g_w),   any(which_loop(g_w)))
) |> kable(caption = "Network type checklist for our toy graph")
```

## R objects for network data

Three representations are used across this workshop:

| Object | Package | When to use |
|---|---|---|
| `igraph` | igraph | Most analyses — fast C backend |
| `tbl_graph` | tidygraph | When you need `dplyr`/`ggraph` pipelines |
| `matrix` / `tibble` | base / tidyverse | Raw import and pre-processing |

```{r}
#| label: r-objects

# igraph object
g_ig <- graph_from_data_frame(el_w, directed = TRUE)

# tbl_graph — wraps igraph, activates nodes or edges as a tibble
g_tbl <- as_tbl_graph(g_ig)

# inspect
g_tbl
```

> **Exercise 0**: Make the toy graph undirected (`as.undirected(g_ig)`). Check again: is the adjacency matrix now symmetric? Does the edge count change, and why?

---

# Part 1 · Graph Models `[~55 min]`

> **Goal**: Understand the three canonical random graph models as null hypotheses for real biological networks.

## Erdős–Rényi: the null model

The ER model places edges between *N* nodes with probability *p*, independently. It predicts a Poisson degree distribution — no hubs. Real biological networks almost never look like this, which is exactly why it is useful as a null.

```{r}
#| label: er-phase-transition
#| fig-cap: "Phase transition in ER networks: a giant component suddenly appears at ⟨k⟩ = 1"

set.seed(42)
n       <- 500
probs   <- seq(0, 0.012, length.out = 80)

er_stats <- map_dfr(probs, function(p) {
  g     <- sample_gnp(n, p, directed = FALSE)
  comps <- components(g)
  tibble(mean_k     = mean(degree(g)),
         giant_frac = max(comps$csize) / n)
})

ggplot(er_stats, aes(x = mean_k, y = giant_frac)) +
  geom_point(colour = "steelblue", size = 1.5) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "firebrick") +
  labs(x = "Mean degree ⟨k⟩", y = "Giant component / N",
       title = "ER percolation threshold",
       caption = "Dashed line: ⟨k⟩ = 1 — the point where a giant component emerges") +
  theme_net()
```

> **Q**: A pathogen progressively disables random host proteins. Which side of this threshold matters for the host? For the pathogen's spread through a PPI network?

## Barabási–Albert: preferential attachment

Each new node attaches with probability proportional to existing degree — "rich get richer". This produces a power-law degree distribution and explains why hubs appear in PPIs, metabolic networks, and co-authorship graphs [@Barabasi1999a].

```{r}
#| label: ba-degree
#| fig-cap: "BA model: straight line on log-log axes is the signature of a power law"

set.seed(42)
ba <- sample_pa(2000, power = 1, directed = FALSE)

tibble(degree = degree(ba)) |>
  count(degree) |>
  ggplot(aes(x = degree, y = n)) +
  geom_point(colour = "coral2") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Degree (log)", y = "Count (log)",
       title = "BA model degree distribution") +
  theme_net()
```

## Watts–Strogatz: small-world networks

A regular lattice rewired with probability β. At intermediate β: short average path lengths (like a random graph) AND high clustering (like a lattice) — simultaneously. This is the *small-world* regime [@Watts1998].

```{r}
#| label: sw-regime
#| fig-cap: "Small-world window: C stays high while L drops, between β ≈ 0.01 and β ≈ 0.1"

set.seed(42)
betas   <- 10^seq(-3, 0, length.out = 40)
lattice <- make_lattice(length = 200, dim = 1, nei = 4, circular = TRUE)
C0      <- transitivity(lattice, type = "average")
L0      <- mean_distance(lattice, directed = FALSE)

sw_stats <- map_dfr(betas, function(b) {
  g <- sample_smallworld(1, 200, 4, b)
  tibble(beta   = b,
         C_norm = transitivity(g, type = "average") / C0,
         L_norm = mean_distance(g, directed = FALSE)  / L0)
})

sw_stats |>
  pivot_longer(c(C_norm, L_norm), names_to = "metric", values_to = "value") |>
  ggplot(aes(x = beta, y = value, colour = metric)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +
  scale_colour_manual(values   = c(C_norm = "steelblue", L_norm = "coral2"),
                      labels   = c("C / C₀  (clustering)", "L / L₀  (path length)")) +
  labs(x = "Rewiring probability β (log)", y = "Normalised value",
       colour = NULL, title = "Watts–Strogatz small-world regime") +
  theme_net()
```

> **Exercise 1**: Generate ER and BA networks each with N = 300 and approximately the same number of edges as a Watts–Strogatz graph with N = 300, K = 4, β = 0.05. Compare clustering coefficient and mean path length across the three. Which model is "most small-world"?

---

# Part 2 · C. elegans Connectome `[~65 min]`

The frontal neuronal network of *C. elegans*: 302 neurons (nodes), chemical synapses and gap junctions (directed edges). Fully mapped from electron microscopy — one of the few complete connectomes in biology [@Kaiser2006].

## Import and network checklist

Every network analysis should start with this checklist before computing anything.

```{r}
#| label: celegans-import

edges <- read_delim("../Data/C-elegans-frontal.txt", delim = " ",
                    col_types = "cci",
                    show_col_types = FALSE) |>
  mutate(across(c(FromNodeId, ToNodeId), as.character))

meta  <- read_csv("../Data/C-elegans-frontal-meta.csv",
                  col_types = cols(), show_col_types = FALSE) |>
  rename(x = posx, y = posy) |>
  mutate(node_id = as.character(node_id))

net <- graph_from_data_frame(edges, directed = TRUE, vertices = meta)

tibble(
  property = c("Nodes", "Edges", "Directed", "Simple",
               "Weighted", "Connected", "Components"),
  value    = c(vcount(net), ecount(net), is_directed(net),
               is_simple(net), is_weighted(net), is_connected(net),
               components(net)$no)
) |> kable(caption = "C. elegans connectome — network checklist")
```

```{r}
#| label: giant-component

# Work with the largest weakly connected component
giant <- decompose(net, min.vertices = 10)[[1]]
cat("Giant component:", vcount(giant), "nodes |", ecount(giant), "edges\n")
```

## Spatial visualisation

Neuron positions are anatomical (inferred from EM data), so this layout is biologically meaningful — not a force-directed aesthetic choice.

```{r}
#| label: spatial-plot
#| fig-cap: "C. elegans frontal connectome in anatomical space"

tidy_net <- as_tbl_graph(giant)

ggraph(tidy_net, layout = "manual", x = x, y = y) +
  geom_edge_link(alpha = 0.2, edge_width = 0.3,
                 arrow     = arrow(length = unit(2, "mm"), type = "closed"),
                 end_cap   = circle(2, "mm")) +
  geom_node_point(size = 2, colour = "steelblue") +
  labs(title = "C. elegans frontal connectome — anatomical layout") +
  theme_void()
```

## Degree distribution and power-law test

```{r}
#| label: degree-powerlaw
#| fig-cap: "Degree distribution: empirical (blue) vs power-law fit (red line from xmin)"

deg_all <- degree(giant, mode = "total")
deg_pos <- deg_all[deg_all > 0]

# Maximum-likelihood fit following Clauset, Shalizi & Newman (2009)
pl      <- displ$new(deg_pos)
pl$setXmin(estimate_xmin(pl))
pl$setPars(estimate_pars(pl))

cat(sprintf("Power-law: α = %.3f  |  xmin = %d  |  tail n = %d\n",
            pl$pars, pl$xmin, sum(deg_pos >= pl$xmin)))

# Empirical complementary CDF
ccdf_df <- tibble(k = sort(unique(deg_pos))) |>
  mutate(p = map_dbl(k, \(x) mean(deg_pos >= x)))

# Theoretical power-law CCDF
k_range     <- seq(pl$xmin, max(deg_pos))
tail_frac   <- mean(deg_pos >= pl$xmin)
pl_ccdf_df  <- tibble(
  k = k_range,
  p = tail_frac * (k_range / pl$xmin) ^ (-(pl$pars - 1))
)

ggplot(ccdf_df, aes(x = k, y = p)) +
  geom_point(colour = "steelblue", size = 1.5) +
  geom_line(data = pl_ccdf_df, aes(x = k, y = p),
            colour = "firebrick", linewidth = 1) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Degree k (log)", y = "P(K ≥ k)  [log]",
       title = "C. elegans degree distribution",
       subtitle = sprintf("Power-law fit: α = %.2f, xmin = %d", pl$pars, pl$xmin)) +
  theme_net()
```

## Centralities

```{r}
#| label: centralities

tidy_net <- tidy_net |>
  activate(nodes) |>
  mutate(
    degree      = centrality_degree(mode = "total"),
    degree_in   = centrality_degree(mode = "in"),
    degree_out  = centrality_degree(mode = "out"),
    betweenness = centrality_betweenness(directed = TRUE),
    closeness   = centrality_closeness(mode = "all"),
    pagerank    = centrality_pagerank()
  )

node_tbl <- tidy_net |> activate(nodes) |> as_tibble()
```

Degree measures local connectivity. Betweenness measures how often a neuron lies on the shortest path between other pairs — a bottleneck. PageRank propagates influence through the directed structure.

```{r}
#| label: centrality-scatter
#| fig-cap: "Nodes above the trend are bottlenecks: high betweenness despite moderate degree"

ggplot(node_tbl, aes(x = degree, y = betweenness)) +
  geom_point(colour = "steelblue", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick", linewidth = 0.8) +
  labs(x = "Degree", y = "Betweenness centrality",
       title = "Hubs vs bottlenecks in C. elegans") +
  theme_net()
```

```{r}
#| label: top-neurons

node_tbl |>
  arrange(desc(betweenness)) |>
  select(name, degree, betweenness, closeness, pagerank) |>
  head(10) |>
  kable(digits = 3,
        caption = "Top-10 neurons by betweenness centrality",
        col.names = c("Neuron", "Degree", "Betweenness", "Closeness", "PageRank"))
```

> **Exercise 2**: Identify neurons in the top-10 by betweenness that are *not* in the top-10 by degree. These are the biological bottlenecks. Look up one of them — what is its known function in *C. elegans* neurobiology?

---

# Part 3 · Community Structure `[~55 min]`

Neurons do not operate in isolation — they form functional modules. Community detection asks: which sets of neurons are more densely interconnected with each other than with the rest?

## Leiden algorithm

Leiden [@Traag2019] solves a resolution-limit problem in Louvain: it guarantees that all communities are internally well-connected and produces more stable partitions. It is the current standard for biological network analysis.

```{r}
#| label: leiden-communities

# Leiden requires an undirected simple graph
tidy_ud <- tidy_net |>
  convert(to_undirected) |>
  convert(to_simple) |>
  mutate(community = as.factor(group_leiden(resolution = 0.5)))

n_comm <- tidy_ud |> activate(nodes) |> pull(community) |> n_distinct()
cat("Communities detected (resolution = 0.5):", n_comm, "\n")
```

```{r}
#| label: spatial-communities
#| fig-cap: "Community membership mapped onto anatomical positions — do communities cluster spatially?"

ggraph(tidy_ud, layout = "manual", x = x, y = y) +
  geom_edge_link(alpha = 0.12, edge_width = 0.3) +
  geom_node_point(aes(colour = community), size = 3) +
  labs(title  = "C. elegans communities (Leiden, resolution = 0.5)",
       colour = "Community") +
  theme_void() +
  theme(legend.position = "bottom")
```

```{r}
#| label: kk-communities
#| fig-cap: "Force-directed layout reveals intra-community density regardless of anatomy"

ggraph(tidy_ud, layout = "kk") +
  geom_edge_link(alpha = 0.12, edge_width = 0.3) +
  geom_node_point(aes(colour = community), size = 3) +
  labs(title = "C. elegans communities (force-directed layout)") +
  theme_void() +
  theme(legend.position = "bottom")
```

## Resolution sensitivity

The `resolution` parameter controls granularity: higher values split communities further. There is no universally correct value — choose based on the biological question.

```{r}
#| label: resolution-sweep
#| fig-cap: "Number of detected communities increases with resolution parameter"

resolutions <- seq(0.1, 2.5, by = 0.1)

res_df <- map_dfr(resolutions, function(r) {
  n <- tidy_ud |>
    mutate(community = group_leiden(resolution = r)) |>
    activate(nodes) |>
    pull(community) |>
    n_distinct()
  tibble(resolution = r, n_communities = n)
})

ggplot(res_df, aes(x = resolution, y = n_communities)) +
  geom_line(colour = "steelblue") +
  geom_point(colour = "steelblue") +
  labs(x = "Resolution parameter", y = "Number of communities",
       title = "Leiden resolution sensitivity") +
  theme_net()
```

## Degree assortativity

Do highly connected neurons preferentially synapse onto other highly connected neurons?

```{r}
#| label: assortativity

cat("Degree assortativity:", round(assortativity_degree(giant, directed = TRUE), 3), "\n")

knn_vals <- knn(giant)$knn
tibble(name = names(knn_vals), knn = knn_vals) |>
  left_join(node_tbl, by = "name") |>
  ggplot(aes(x = degree, y = knn)) +
  geom_point(colour = "steelblue", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Degree", y = "Mean neighbour degree",
       title = "Degree assortativity in C. elegans") +
  theme_net()
```

> **Exercise 3**: Change the Leiden `resolution` to 1.5. Do the communities still align with anatomical positions, or do they fragment? What does a negative assortativity value mean biologically in a neural circuit?

---

# Part 4 · E. coli PPI and Network Robustness `[~50 min]`

## Protein–protein interaction network

BioGRID curates experimentally validated PPIs from the literature [@Stark2006]. We keep only *physical* interactions (direct binding or stable complex), not *genetic* interactions (functional relationships without physical contact).

```{r}
#| label: ecoli-import

biogrid_raw <- read.delim(
  "../Data/BIOGRID-ORGANISM-Escherichia_coli_K12_W3110-3.5.165.mitab.txt",
  sep = "\t", header = TRUE
)

physical_types <- c(
  "psi-mi:MI:0407(direct interaction)",
  "psi-mi:MI:0915(physical association)"
)

ppi_edges <- biogrid_raw |>
  filter(Interaction.Types %in% physical_types) |>
  select(from = Alt.IDs.Interactor.A, to = Alt.IDs.Interactor.B)

ppi_net <- graph_from_data_frame(ppi_edges, directed = FALSE) |>
  simplify()

cat("E. coli PPI — nodes:", vcount(ppi_net),
    "| edges:", ecount(ppi_net),
    "| density:", round(edge_density(ppi_net), 5), "\n")
```

```{r}
#| label: ecoli-degree
#| fig-cap: "E. coli PPI degree distribution — linear on log-log, consistent with a power law"

deg_ppi <- degree(ppi_net)

tibble(degree = deg_ppi) |>
  count(degree) |>
  ggplot(aes(x = degree, y = n)) +
  geom_point(colour = "coral2") +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Degree (log)", y = "Count (log)",
       title = "E. coli PPI degree distribution") +
  theme_net()
```

## Robustness: random failures vs targeted attacks

Scale-free networks tolerate random node removal (rare hubs are unlikely to be hit) but collapse quickly under targeted hub removal. This asymmetry has direct implications for antibiotic target selection and network-based drug design.

```{r}
#| label: robustness
#| cache: true
#| fig-cap: "Percolation: the PPI network is robust to random failures but fragile under hub removal"

percolate <- function(g, strategy = c("random", "targeted"), n_steps = 80) {
  strategy <- match.arg(strategy)
  N        <- vcount(g)
  V(g)$id  <- seq_len(N)

  step_at <- round(seq(1, N - 1, length.out = n_steps))

  results <- vector("list", n_steps)
  for (i in seq_along(step_at)) {
    comps <- components(g)
    results[[i]] <- tibble(
      removed_frac = (N - vcount(g)) / N,
      giant_frac   = max(comps$csize)  / N
    )
    del <- if (strategy == "random") {
      sample(V(g)$id, 1)
    } else {
      which.max(degree(g))
    }
    g <- delete_vertices(g, del)
    V(g)$id <- seq_len(vcount(g))
  }
  bind_rows(results) |> mutate(strategy = strategy)
}

set.seed(42)
ppi_giant  <- decompose(ppi_net)[[1]]
robust_df  <- bind_rows(
  percolate(ppi_giant, "random"),
  percolate(ppi_giant, "targeted")
)

ggplot(robust_df, aes(x = removed_frac, y = giant_frac, colour = strategy)) +
  geom_line(linewidth = 1) +
  scale_colour_manual(values = c(random = "steelblue", targeted = "firebrick"),
                      labels = c("Random failure", "Targeted hub removal")) +
  labs(x = "Fraction of proteins removed",
       y = "Giant component fraction",
       colour = NULL,
       title  = "E. coli PPI robustness") +
  theme_net()
```

## Network inference: from correlation to edges

Most biological networks are not directly observed — they are *inferred* from expression data. The simplest approach: correlate gene expression profiles across conditions and apply a threshold. The threshold choice is a biological decision, not a statistical one.

```{r}
#| label: inference-threshold
#| fig-cap: "Network structure is highly sensitive to the correlation threshold — there is no free lunch"

set.seed(42)
n_genes   <- 60
n_samples <- 25
expr      <- matrix(rnorm(n_genes * n_samples), nrow = n_genes,
                    dimnames = list(paste0("g", seq_len(n_genes)), NULL))

# Plant a co-expression module in genes 1–15
expr[1:15, ] <- expr[1:15, ] + rnorm(n_samples, mean = 2.5, sd = 0.5)

cor_mat    <- cor(t(expr))
thresholds <- seq(0.3, 0.9, by = 0.05)

thr_stats <- map_dfr(thresholds, function(thr) {
  adj       <- (abs(cor_mat) > thr) * 1L
  diag(adj) <- 0L
  g         <- graph_from_adjacency_matrix(adj, mode = "undirected")
  tibble(threshold    = thr,
         density      = edge_density(g),
         n_components = components(g)$no,
         mean_degree  = mean(degree(g)))
})

thr_stats |>
  pivot_longer(-threshold, names_to = "metric", values_to = "value") |>
  ggplot(aes(x = threshold, y = value)) +
  geom_line(colour = "steelblue") + geom_point(colour = "steelblue") +
  facet_wrap(~ metric, scales = "free_y",
             labeller = as_labeller(c(
               density      = "Edge density",
               n_components = "Connected components",
               mean_degree  = "Mean degree"
             ))) +
  labs(x = "Correlation threshold", y = NULL,
       title = "Effect of threshold on inferred network structure") +
  theme_net()
```

> **Exercise 4**: At what threshold does the inferred network fragment into more than 10 components? Is a denser or sparser network more biologically credible for a 25-sample experiment? What additional information (besides the threshold) would you need to validate this network?

---

# Part 5 · Networks as Computation `[~40 min]`

> **Goal**: Recognise deep learning models as directed weighted graphs and apply the same structural questions we asked all day — degree, density, topology — to artificial neural networks and transformers.

## From biological to artificial neurons

The graph abstraction does not only *describe* biological systems — it is also the computational substrate of modern AI. An artificial neural network is a directed weighted graph:

- **Nodes** are artificial neurons (units that apply a non-linear function to their inputs).
- **Edges** are synaptic weights (learned real numbers).
- **Layers** impose a topological constraint: edges only go forward, from layer $l$ to layer $l+1$.

The McCulloch–Pitts neuron (1943) was the first formal model. The perceptron (Rosenblatt, 1958) added learning. The backpropagation algorithm (Rumelhart et al., 1986) made deep multi-layer networks trainable. Transformers [@Vaswani2017] (2017) replaced fixed topology with *dynamic* edge weights — the attention mechanism — and became the architecture behind large language models (LLMs).

The key conceptual bridge: **every technique we used today applies equally to biological and artificial networks**. The questions change; the tools do not.

## Deep feedforward network topology

A **multi-layer perceptron** (MLP) is a fully connected layered graph. We build one from scratch as an igraph object and ask the same structural questions we asked of the *C. elegans* connectome.

```{r}
#| label: mlp-build

layer_sizes <- c(input = 4, hidden1 = 6, hidden2 = 6, output = 2)

# Node table: one row per neuron
nodes_mlp <- imap_dfr(layer_sizes, function(n, lname) {
  tibble(
    layer      = lname,
    layer_idx  = which(names(layer_sizes) == lname),
    neuron_idx = seq_len(n),
    name       = paste0(lname, "_", seq_len(n))
  )
}) |> mutate(
  x = layer_idx,
  y = neuron_idx - (layer_sizes[layer] + 1) / 2   # centre each layer vertically
)

# Edge table: every neuron in layer l → every neuron in layer l+1
layer_names <- names(layer_sizes)
edges_mlp <- map_dfr(seq_len(length(layer_sizes) - 1), function(i) {
  from_nodes <- nodes_mlp |> filter(layer == layer_names[i])
  to_nodes   <- nodes_mlp |> filter(layer == layer_names[i + 1])
  crossing(from = from_nodes$name, to = to_nodes$name)
})

g_mlp <- graph_from_data_frame(edges_mlp, vertices = select(nodes_mlp, name, everything()), directed = TRUE)

tibble(
  Property = c("Neurons (nodes)", "Weights (edges)", "Layers",
               "Density", "Is DAG"),
  Value    = c(vcount(g_mlp), ecount(g_mlp), length(layer_sizes),
               round(edge_density(g_mlp), 3),
               is_dag(g_mlp))
) |> kable(caption = "MLP network checklist")
```

```{r}
#| label: mlp-plot
#| fig-cap: "MLP as a directed graph — each column is a layer, each node a neuron, each edge a learnable weight"

as_tbl_graph(g_mlp) |>
  activate(nodes) |>
  mutate(layer = factor(layer, levels = layer_names)) |>
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(alpha = 0.25, edge_width = 0.4,
                 arrow   = arrow(length = unit(2, "mm"), type = "closed"),
                 end_cap = circle(3, "mm")) +
  geom_node_point(aes(colour = layer), size = 6) +
  geom_node_text(aes(label = neuron_idx), size = 3, colour = "white") +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Multi-layer perceptron — fully connected", colour = "Layer") +
  theme_void()
```

The degree distribution of a fully connected MLP is nothing like a biological network: every neuron has *exactly* the same degree within each layer — a perfectly regular graph with no hubs.

```{r}
#| label: mlp-degree
#| fig-cap: "MLP degree distribution: a spike — every node has the same degree"

tibble(degree = degree(g_mlp, mode = "total")) |>
  count(degree) |>
  ggplot(aes(x = degree, y = n)) +
  geom_col(fill = "steelblue", width = 0.6) +
  labs(x = "Degree", y = "Count",
       title = "MLP degree distribution — no hubs, no tail") +
  theme_net()
```

## Scale: from a toy MLP to GPT

The parameter count of a real model dwarfs anything in biology. How does that translate to graph terms?

```{r}
#| label: scale-table

params <- function(layers) {
  # weights + biases between consecutive layers
  n <- length(layers)
  sum(layers[-n] * layers[-1]) + sum(layers[-1])
}

scale_df <- tribble(
  ~Model,                ~Nodes,       ~Edges_weights,   ~Note,
  "C. elegans (frontal)", "131",        "764",            "measured connectome",
  "E. coli PPI",          "~2 400",     "~5 200",         "BioGRID physical",
  "MLP 4-6-6-2",          "18",         as.character(params(c(4,6,6,2))),   "this workshop",
  "MLP 784-256-128-10",   "1 178",      as.character(params(c(784,256,128,10))), "MNIST classifier",
  "ResNet-50",            "~25 M",      "~25 M",          "image classifier (2015)",
  "GPT-2 small",          "~124 M",     "~124 M",         "language model (2019)",
  "GPT-4 (estimated)",    "~1.8 T",     "~1.8 T",         "language model (2023)"
)

kable(scale_df,
      col.names = c("Model", "Nodes / Neurons", "Edges / Parameters", "Context"),
      caption   = "Networks at scale: biological and artificial")
```

> **Q**: The *C. elegans* connectome has ~130 neurons. GPT-2 has ~124 million parameters. What does this difference imply about the information-processing strategies of biological vs artificial systems?

## Transformer attention as a dynamic network

In a transformer, attention replaces the fixed topology of an MLP with a **learned, input-dependent weighted graph** over the sequence. For each input (e.g., a sentence), the model computes a weight matrix — how much each token "attends to" every other token. This is a complete directed weighted graph whose edges change with every input.

Simulate one attention head over a 12-token sequence:

```{r}
#| label: attention-simulate

set.seed(42)
tokens    <- c("The", "cat", "sat", "on", "the", "mat", "and", "ate", "a", "rat", ".", "[EOS]")
n_tok     <- length(tokens)

# Raw attention scores (dot-product of query and key vectors, simulated)
logits    <- matrix(rnorm(n_tok^2, sd = 1.5), n_tok, n_tok)

# Causal mask: tokens cannot attend to future tokens (GPT-style)
mask           <- matrix(0, n_tok, n_tok)
mask[upper.tri(mask, diag = FALSE)] <- -Inf
logits_masked  <- logits + mask

# Softmax row-wise → attention weights (rows sum to 1)
softmax_row <- function(x) { e <- exp(x - max(x[is.finite(x)])); e / sum(e) }
attn        <- t(apply(logits_masked, 1, softmax_row))
rownames(attn) <- colnames(attn) <- tokens
```

```{r}
#| label: attention-heatmap
#| fig-cap: "Attention matrix: rows are queries (what attends), columns are keys (what is attended to). Upper triangle is masked — GPT-style causal attention."

as.data.frame(attn) |>
  rownames_to_column("query") |>
  pivot_longer(-query, names_to = "key", values_to = "weight") |>
  mutate(
    query = factor(query, levels = rev(tokens)),
    key   = factor(key,   levels = tokens)
  ) |>
  ggplot(aes(x = key, y = query, fill = weight)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  scale_fill_gradient(low = "grey95", high = "steelblue",
                      na.value = "grey80", name = "Attention\nweight") +
  labs(x = "Key (attended to)", y = "Query (attending)",
       title = "Transformer self-attention — one head") +
  theme_net() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Convert the attention matrix to a sparse graph by keeping only the top weights:

```{r}
#| label: attention-graph
#| fig-cap: "Attention graph: directed edges connect tokens to the tokens they attend most strongly (top 20% of weights)"

threshold <- quantile(attn[is.finite(attn)], 0.80)

attn_adj           <- (attn >= threshold) * 1L
attn_adj[!is.finite(attn)] <- 0L

g_attn <- graph_from_adjacency_matrix(attn_adj, mode = "directed", weighted = NULL)

tibble(
  Property = c("Tokens (nodes)", "Attention edges", "Density",
               "Mean in-degree", "Max in-degree token"),
  Value    = c(vcount(g_attn), ecount(g_attn),
               round(edge_density(g_attn), 3),
               round(mean(degree(g_attn, mode = "in")), 2),
               tokens[which.max(degree(g_attn, mode = "in"))])
) |> kable(caption = "Attention graph — structural summary")
```

```{r}
#| label: attention-network-plot
#| fig-cap: "Attention as a network: node size = in-degree (how often a token is attended to)"

as_tbl_graph(g_attn) |>
  activate(nodes) |>
  mutate(indegree = centrality_degree(mode = "in")) |>
  ggraph(layout = "circle") +
  geom_edge_arc(alpha = 0.4, edge_width = 0.5,
                arrow   = arrow(length = unit(2, "mm"), type = "closed"),
                end_cap = circle(4, "mm"),
                strength = 0.2) +
  geom_node_point(aes(size = indegree), colour = "steelblue") +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_size_continuous(range = c(2, 8)) +
  labs(title = "Attention graph (top 20% weights)", size = "In-degree") +
  theme_void()
```

## What makes artificial and biological networks different?

```{r}
#| label: comparison-table

tribble(
  ~Property,           ~`C. elegans`,           ~`MLP`,                        ~`Transformer`,
  "Topology",          "Irregular, sparse",      "Layered, fully connected",    "Dynamic — changes per input",
  "Degree dist.",      "Heterogeneous (hubs)",   "Regular (all equal)",         "Input-dependent",
  "Edge weights",      "Discrete (synapse count)","Continuous (learned reals)", "Continuous (attention scores)",
  "Directionality",    "Directed",               "Directed (forward pass)",     "Directed (causal mask)",
  "Learning",          "Synaptic plasticity",    "Gradient descent",            "Gradient descent",
  "Scale (edges)",     "~800",                   "~100 – 100 M",               "~100 M – 1 T"
) |> kable(caption = "Biological vs artificial networks — structural comparison")
```

> **Exercise 5**: The attention heatmap shows which tokens attend to which. Look at the token with the highest in-degree. Does it make linguistic sense that it would be widely attended to? Now change `set.seed(42)` to another value — does the same token remain the hub, or does the result change? What does this tell you about interpreting attention weights as meaningful connections?

---

# References {.unnumbered}
